Uniform formula lower bounds for finding and counting satisfiable assignments

### 2022-05-18 article/main.tex
Added todo

@@

%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%
\title{Alternating formula lower bounds for SAT}
\author{Gabriel Bathie, Ryan Williams}
\begin{document}
\maketitle

@@

List of things to do:
\begin{itemize}
	\item Improve Spira/Brent to show that a $c \log n$depth 
	lower bound implies an $n^{c-\eps}$ size lower bound?
	\item Show that this bound is optimal w.r.t. to these rules?
\end{itemize}

@@

Motivation

@@

Main results:
\begin{theorem}
	For any $c < 4 \cos(\pi/7)$, \SAT does not have \NAND formulas of depth $(c + o(1)) \log n$.
\end{theorem}

\todo[inline]{Prove all the following}
\begin{lemma}
	For any \NAND formula on $n$ inputs of size $n^c$ and any $\eps > 0$, 
	there exists an equivalent \NAND formula of depth $(c+\eps) \log n$.
\end{lemma}

\begin{corollary}
	For any $c < 4 \cos(\pi/7)$, \SAT does not have \NAND formulas of \textbf{size} $n^{c + o(1)}$.
\end{corollary}

@@

Uniform \NAND/alternating formulas. Specific model, comparison with the usual model.

Alternating computation.

Approach: proof by contradiction.

@@

\subsection{Improving Brent/Spira}
\todo[inline]{Prove this}

%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%
\section{Proof of optimality}
\todo[inline]{Prove this}

>>>>

\newcommand{\Omegat}[1]{\widetilde{\Omega}\left( #1 \right)}%
\newcommand{\Ot}[1]{\widetilde{O}\left( #1 \right)}%

@@

%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%
\title{Uniform formula lower bounds for finding and counting satisfiable assignments}
\author{Gabriel Bathie, Ryan Williams}
\begin{document}
\maketitle

@@

List of things to do:
\begin{itemize}
	\item Improve Spira/Brent to show that a $c \log n$ depth 
	lower bound implies an $n^{c-\eps}$ size lower bound?
	\item Show that this bound is optimal w.r.t. these rules?
\end{itemize}

@@

Motivation: understanding complexity. Explicit lower bounds : programs, circuits, formulas.
Uniformity.

@@

Best known formula lower bounds for languages (functions?) in NP: 
\begin{itemize}
	\item $\Omegat{n^3}$ (for both uniform and non-uniform formulas) for Andreev's function \cite{hastad1998shrinkage}\todo{check source} in the De Morgan basis $\{\wedge, \vee, \neg\}$.
	\item $\Omega{n^2/\log n}$ in the full binary basis (Nechiporuk)
	\item $n^{\Theta(\log n)}$ for monotone formulas (formulas over the $\{\wedge, \vee\}$ basis).
\end{itemize}

\subsection{Results overview}

We first study \NAND{} formulas. 
The \NAND{} basis is rather expressive: Andreev's function has \NAND{} formulas of depth $(3+o(1)) \log n$ (and therefore of size $\Ot{n^3}$), which matches the lower bound for the De Morgan Basis.

\begin{theorem}
	For any $c < 4 \cos(\pi/7) \simeq 3.603$, \SAT does not have \NAND{} formulas of depth $(c + o(1)) \log n$.
\end{theorem}

\todo[inline]{Prove all of the following.

After checking \cite{sergeev2019relation}, this is probably incorrect...}
\begin{lemma}
	For any \NAND{} formula on $n$ inputs of size $n^c$ and any $\eps > 0$, 
	there exists an equivalent \NAND{} formula of depth $(c+\eps) \log n$.
\end{lemma}

\begin{corollary}
	For any $c < 4 \cos(\pi/7)$, \SAT does not have \NAND{} formulas of \textbf{size} $n^{c + o(1)}$.
\end{corollary}

@@

Uniform \NAND{}/alternating formulas. Specific model, comparison with the usual model.

\subsection{Alternating computation}

Given a complexity class $\Cc$ and a constant $c$, we consider \textit{quantified} complexity
class $(Q~n^c)~\Cc$, where $Q$ is one of $\forall, \exists$ or $\oplus$.
We define them as follows:
\begin{definition}[Quantified classes]
	A language $L$ is in $(Q~n^c)~\Cc$ if and only if there exists a RAM machine
	that runs in time $n^{c + o(1)}$...

	Problem with this definition: size of the input $n$ is not the same at every stage.
\end{definition}

\subsection{Overview of the methodology}
Approach: proof by contradiction.

@@

\subsection{Improving Brent/Spira}
\todo[inline]{Prove this}
Useful starting point : Size-Depth Tradeoffs for Boolean Formulae (Bonet, Buss).

Remark: we must focus on the case when the input formula is a \NAND{} formula.
Otherwise, we prove a much stronger result.
Maybe prove the stronger result with a $2+\eps$ factor? Not very useful though.
If we can go below 2, we improve something.
%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%
\section{Proof of optimality}
\todo[inline]{Prove this}
